Density-dependent selection and the limits of relative fitness

### 2018-03-19 manuscript.tex
latest

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness in evolutionary genetics, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp. 26]{crow_1970}. In the particular case that two types are present, the selection equation for the first type takes the familiar form
\begin{equation}
\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
\end{equation}
where $s=r_1-r_2$ is the selection coefficient. Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to $r_1$ and $r_2$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{nagylaki_1979,travis_2013}. Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not $N$ as well.

In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. However, there are reasons to doubt that this setup is ecologically valid. Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

>>>>

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp. 26]{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficiento of type $i$ relative to type $j$. Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection still only requires us to model the type frequencies, not $N$ as well.

and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.

A major focus of the literature on density-regulated population growth focuses on the connection between selection and the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.

This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps,

In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability, but other important population ecological assumptions --- such as how density is regulated --- are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).

 It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
